% Chapter 3

\chapter{Image Segmentation}

\lhead{Chapter 3. \emph{Image Segmentation}} % This is for the header on each page - perhaps a shortened title
\label{chp:segmentation}
%----------------------------------------------------------------------------------------

\def\:{\hskip0pt} %Definisce un modo veloce per permettere a latex di sillabare correttamente anche parole come 4-connectivity. Il corretto utilizzo è il seguente: 4\:-\:connectivity.

\section{Introduction}
Image segmentation is a well known topic in computer and image processing with a
wide range of application, such as medical imaging, robotics, video
surveillance, etc.
It involves partitioning images into one or more objects and can also includes
classify these objects. Many traditional algorithms have been developed in the
literature but, in the most recent years, they have all been dominated by deep
neural networks. Since the 2015 a huge amount of different types of networks
that aim to perform image segmentation has been proposed for each of the field
where it's needed.
Before presenting how nowday segmentation is performed, we must state which are
the different type of segmentation that have been classified.

\subsection{Semantic Segmentation}
Semantic Segmentation perform a pixel-by-pixel classification with a predefined
set of objects categories for all the pixels of the images. In pratice, given a
RGB image (\texttt{height} $\times$ \texttt{width} $times$ \texttt{3}) we output
a segmentation map of size (\texttt{height} $\times$ \texttt{width} $\times$
\texttt{1}) where each value correspond to which class the same pixel in the
original images belongs.

\subsection{Instance segmentation}
One possible issue of semantic segmentation is that it doesn't allow to
distinguish two or more object of the same class when they overlap in the image.
Instance segmentation overcome this problem by outputting a different number of
channels based on the number of instances present in the image.

\subsection{Panoptic segmentation}
The latter type of segmentation is called Panoptic segmentation and is the
result of the previously presented method joined together. The difference with
the instance segmentation is that in this case instances are not allowed to
overlap then for to a single pixel, a single instance must be assigned.

\subsection{Segmentation using Deep Neural Networks}
As stated before, nowdays deep learning is the dominant method used for
segmentation. Convolutional Neural Networks are still the most used type of
networks used to date but new promising technique, as Graph Neural Networks or
Transformers, are always more and more used.
